Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-08-18 thesis/introduction.tex
thesis julio ch1 edits

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These little galaxies are typically
defined to be fainter than the LMC \citep[\(M_V \gtrsim -18\) or
similarly \(L \lesssim 10^9 L_\odot\),
e.g.,][]{hodge1971, mcconnachie2012}. As the smallest class of galaxies,
dwarfs are the most cosmologically numerous galaxies. Dwarf galaxies are
highly \emph{dark-matter dominated}, with mass to light ratios exceeding
100 to 1000 \(M_\odot/ L_\odot\). Because of their small total masses,
many dwarf galaxies are \emph{quenched}, with little to no recent star
formation. Their stellar populations are \emph{relics} from the early
universe, consisting of many of the oldest and most metal poor stars.
Understanding the properties of dwarf galaxies thus has implications
across astronomy, from cosmological structure formation on the smallest
scale to extremely metal-poor stellar populations to chemical evolution
to galactic dynamics. Of the classical dwarfs, the Sculptor and Ursa
Minor dwarf spheroidal galaxies stand out with a more extended density
profile relative to an exponential, hinting at tidal effects. As a case
study in tidal effects and our understanding of dwarf galaxy formation
and evolution, we aim to understand the dynamical history of these
galaxies.

@@

We only understand a tiny fraction of the universe's composition. The
leading theory of cosmology, \(\Lambda\)CDM (cold dark matter), posits
that the universe is composed of about 68\% dark energy (\(\Lambda\)),
27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers like
  to change definitions of words. \emph{Baryons} here means
  baryons+leptons, i.e.~any standard model massive fermion. The photon
  energy density is negligible today} \citep{planckcollaboration+2020}.
While the composition of dark matter and dark energy remains elusive, we
know their general properties. Dark energy causes the acceleration of
the expansion of the universe on large scales. We do not discuss dark
energy here---it does not substantially affect the local group today.
Dark matter, instead, makes up the vast majority of mass in galaxies.
Typically, galaxies have baryonic to dark matter ratios of between 1:5
to beyond 1:1000 for faint dwarf galaxies. In \(\Lambda\)CMD, dark
matter is assumed to only interact gravitationally, passing through
matter without effect (transparent to light, or \emph{dark}). Dark
matter is also \emph{cold}, i.e.~typical velocities much smaller than
the speed of light in the early universe. Implications of dark matter
properties range from cosmological structural formation, galaxy
structure, and galaxy interactions.

The very early universe was almost featureless. Our earliest
observations of the universe stem from the cosmic microwave background
(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
But tiny perturbations in the CMB, temperature fluctuations of 1 part in
10,000, became the seeds of large-scale cosmological structure. Governed
by cosmological expansion, gravitational collapse, and baryonic physics,
each perturbation grows into larger structures. Initially, baryonic
matter was coupled to radiation and resisted collapse. Dark matter, only
influenced by gravity instead, freely collapsed into the first
structures. Each self-gravitating overdensity of dark matter is known as
a \emph{halo}. After recombination, where electrons combined with atomic
nuclei to form atoms, baryons decoupled from radiation and fell into the
dark matter halos. The densest pockets of baryons later formed the first
stars and galaxies.

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, structure formation is \emph{hierarchical}. Small
dark matter halos collapse first and hierarchically merge into
progressively larger halos
\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
Structure formation happens on a wide range of scales. The largest
structures become the cosmic web---composed of voids, filaments, and
clusters---and the smallest structures directly observable host dwarf
galaxies. Hierarchical assembly is evident through the large scale
structure of the universe, remnants of past mergers within the milky
way, and tidal disruption of dwarf galaxies and their streams around
nearby galaxies.

Small-scale structure formation is sensitive to deviations from
\(\Lambda\)CDM cosmology \citep[e.g.,][]{bechtol+2022}. Many alternative
models, such as warm dark matter or self-interacting dark matter, may
smooth out small-scale features and reduce the abundance of small halos
or change their structure \citep[e.g.,][]{lovell+2014}. Dwarf galaxies,
occupying the smallest dark matter halos, are promising windows into
small-scale cosmological features. Alternative dark matter properties
can influence dwarf galaxy abundance, formation, structure,, and tidal
evolution. The nearby dwarf galaxies of the local group, with detailed
observations, present a promising opportunity to understand the
evolution of these objects and test our understanding of cosmology. To
understand the evolution of these objects, we need to understand the
general predictions from \(\Lambda\)CDM for the properties of dwarf
galaxies.

In \(\Lambda\)CDM cosmological simulations, dark matter halos are
remarkably self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the
authors observe that the spherical radial density profiles \(\rho(r)\)
are well described by a two-parameter law: \begin{equation}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density . This profile has shown remarkable success in describing
\(\Lambda\)CMD halos across several orders of magnitude in mass. NFW
profiles are \emph{cuspy}, where the density continuously rises like
\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
density profile increases at \(r \sim r_s\) and at large radii, the
density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
shows an example of a NFW halo and a cored dark matter halo.

The total mass of an NFW profile diverges, so halos are commonly
characterized by an overdensity criterion. The virial mass, \(M_{200}\),
is defined as the mass within a region \(r_{200}\) containing a mean
enclosed density 200 times\footnote{From approximate analytic arguments
  about the mean density of a self-gravitating spherical collapse halo
  in an expanding universe, the expected virial overdensity is \ldots,
  commonly rounded to 200 (REFS).} the critical density of the universe:
\begin{equation}{
M_{200} =(4\pi/3) \ r_{200}^3\ 200\rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z) / 8\pi G
}\end{equation} A standard second parameter is the halo concentration,
\(c=r_{200} / r_s\) describing how the characteristic size of the halo
compares to its virial radius. In this case, the scale density is a
function of \(c\) alone,
\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
\citep{NFW1996}. However, we characterize halos by their circular
velocity profiles. The circular velocity,
\(v_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum of
\(v_{\rm max}\) at radius \(r_{\rm max} = \alpha\,r_s\) where
\(\alpha\approx2.16258\). \(v_{\rm max}\) is related to both the total
halo mass and the observed line of sight velocity disperion, and
\(r_{\rm max}\) relates to the scale radius.

While an NFW profile has two free parameters (\(M_{200}\) and \(c\)),
these are not independent. Smaller dark matter halos often collapse
earlier, when the universe was denser. As a result, small subhalos tend
to be more concentrated \citep[e.g.,][]{NFW1997}. The relationship
between \(M_{200}\) and c, or the mass-concentration relation, how more
massive halos become less concentrated in a predictable way
\citep[e.g.,][]{bullock+2001, ludlow+2016}. The lower panel of
Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration
relation in terms of \(r_{\rm max}\) and \(v_{\rm max}\). While there is
a general trend in concentration with mass, the concentration values
still have substantial scatter. Other parameters such as the halo spin
or shape may add additional variability to present-day distribution of
halos, but this variation is typically expected to be small
\citep{navarro+2010, dicintio+2013, dutton+maccio2014}.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/cosmological_means.pdf}}
\caption[Stellar-mass halo-mass relation]{\textbf{Top} Stellar mass
(top) as a function of maximum circular velocity. The solid line with
the 1-\(\sigma\) shaded region is the relation from \citet{fattahi+2018}
with scatter points simulated central galaxies from APOSTLE in
\citet{fattahi+2018}. \textbf{Bottom:} Radius of maximum circular
velocity \(r_{\rm max}\) as a function of maximum circular velocity. The
solid line with 1-\(\sigma\) shaded region and dashed line are the
relations from \citet{ludlow+2016} for \(z=0\) and \(z=2\) respectively.
\emph{TODO}: add legend / annotations.}\label{fig:smhm}
\end{figure}

Both cosmology and observations find fundamental relationships between
the dynamical or total mass and stellar masses of galaxies. A key
prediction of \(\Lambda\)CDM is the stellar mass halo mass relation,
describing the amount of stellar mass forms in a (sub)halo of a given
size. In particular, the SMHM relation grows especially steep in the
dwarf galaxy regime---many dwarf galaxies are formed in halos of similar
masses. Fig.~\ref{fig:smhm} shows the stellar mass versus
\(v_{\rm max}\) maximum circular velocity (proxy for halo mass) for
apostle dwarfs from \citet{fattahi+2018}. While there is some scatter,
the range of predicted \(v_{\rm max}\) is fairly narrow across
\(\sim 4\) decades in stellar mass. Because lower mass galaxies have
increasingly shallow potential wells, feedback becomes more effective
removing gas more quickly. In addition, re-ionization suppresses late
star formation in the faintest galaxies. Thus, if we know the initial
stellar mass of a dwarf galaxy, we also know, at some level, the
properties of its dark matter halo.

Several challenges complicate a simple SMHM trend including environment,
assembly history, and tidal effects. Additionally, the detailed physics
of galaxy formation can impact the SMHM relationship. By being closer to
a massive host, many dwarfs quench earlier, resulting in lower stellar
masses \citep[e.g.,][]{christensen+2024}. Additionally, the time of
formation (relative to reionization) can influence the resulting stellar
content \citep{kim+2024}. Finally, tides influence both the dark matter
and stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
Consequently, tides may reduce the halo mass more than the stellar mass,
adding additional scatter to the SMHM trend, particularly for low-mass
satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
tides on local group dwarf galaxies is essential to determining where
and how these galaxies formed in a cosmological context.

@@

Dwarf galaxies have long raised conundrums in theories of galaxy
formation. The discovery of Fornax and Sculptor in 1938
\citep{shapley1938}\footnote{technically, the LMC and SMC may be
  classified as dwarf galaxies, but these were likely always known to
  humans at southern latitudes.}, with no known analogues, already
presented such an enigma. Shapley presented these dwarfs as a new type
of \emph{stellar system} resembling the Magellanic clouds and globular
clusters but did not attempt to speculate on the exact nature. While
dwarf galaxies were quickly understood to be galaxies based on the
inferred luminosities and sizes, their nature and formation remained
unclear for decades \citep[e.g.,][]{hodge1971, gallagher+wyse1994}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial dark matter masses. From early determinations of the
velocity dispersion for Sculptor and Ursa Minor dSph
\citep[e.g.,][\citet{aaronson+olszewski1987}]{aaronson1983}, inferred
mass-to-light ratios were at least 10 times the expectation for globular
clusters. While only determined by a few stars initially, these values
have solidified with time with larger and more precise samples
\citep[e.g.,][]{hargreaves+1994}. Subsequently, several theories
attempting to understand the formation and observed properties of these
objects were proposed. Examples include: dwarf galaxies are undergoing
tidal dissolution resulting in extreme mass-to-light ratios
\citep[e.g.,][]{kuhn+miller1989}, presence of massive central black
holes \citep[e.g.,][]{strobel+lake1994}, the formation of dark matter
free out-of-equilibrium ``tidal dwarfs'' from past mergers
\citep[e.g.,][\citet{kroupa1997}]{lynden-bell1982}, or modified theories
of gravity \citep{milgrom1995}. However, consistency with CDM galaxy
formation was also not out of the question
\citep[e.g.,][]{dekel+silk1986}. Since then, we have known that dwarf
galaxies are among the darkest objects in the universe, and
understanding their properties is critical to understanding the
universe.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs galaxies} with magnitudes
\(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
and recent star formation. Fig.~\ref{fig:galaxy_images} shows the LMC as
an example of an irregular (and slightly bright) dwarf galaxy displaying
a bar. \textbf{Classical dwarf} galaxies occupy intermediate
luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these
systems are old, gas-poor, and spheroidal. All Milky Way satellites
discovered before digital sky surveys are classicals. The 12 classical
dwarfs satellites of our galaxy are Sagittarius, Fornax, Leo I,
Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
I, Sextans I, and Crater II.\footnote{While formally the dwarf galaxy
  names we discuss contain'' dwarf spheroidal'' (dSph), e.g.~Sculptor
  dSph, we omit this suffix for brevity.} The \textbf{ultra-faint}s
occupy the very faintest magnitudes (\(-7.7 \lesssim M_V\)). These
galaxies have minuscule stellar masses, tend to be more compact, and are
the darkest known galaxies. Altogether, dwarf galaxies span about 7
orders of magnitude in stellar mass.

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.pdf}
\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Fornax (DES
DR2), Sculptor (DES DR2), and Ursa Minor (UNWISE with Gaia point sources
over-plotted). Each image includes the galaxies luminosity and a 200 pc
scalebar. While the LMC is very prominent in the sky, the classical
dwarf galaxies are not obvious except in terms of star counts
(e.g.,Fig.~\ref{fig:scl_selection}). All (non-LMC) dwarfs occupy the
central \(\sim\)third of the image.}\label{fig:galaxy_images}
\end{figure}

@@

A star's or galaxy's position and velocity are fundamental quantities
for understanding its orbit and origin. Unfortunately, determining
distances and full 3D velocities is challenging. The most direct
measurement of distance, parallax, requires precise tracking of a star's
sky position across a year. And while line-of-sight (LOS) velocities are
easily determined from spectroscopy, tangental velocities, derived from
proper motions, are much more challenging. Typically, measuring proper
motions requires accurate (much less than arcsecond) determinations of
small changes in a star's position over baselines of several years to
decades. The full 6D position and velocity information for stars was
historically only known for a handful of stars.

\emph{Gaia} has redefined astrometry, providing photometry, positions,
proper motions, and parallaxes for over 1 billion stars
\citep{gaialcollaboration+2021}. Launched in 2013, \emph{Gaia} is a
space-based, all-sky survey telescope situated at the Sun-Earth L2
Lagrange point \citep{gaiacollaboration+2016}. While \emph{Gaia}
completed its space-based mission in 2025, two forthcoming data releases
remain. Determining absolute parallax measurement is facilitated by the
observation that stars in different regions of the sky are affected by
parallax motion with different phases. By imagining two regions
separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
tiny changes in relative positions of stars across small and large
angles. Combining measurements from multiple epochs across several
years, an absolute all-sky reference frame is derived from which
parallax and proper motions are derived \citep{gaiacollaboration+2016}.
In addition to astrometry, \emph{Gaia} measures photometry in the wide
\emph{G} band (330-1050nm) and colours from the blue photometer (BP,
330-680 nm) and red photometer (RP, 640-1050 nm). \emph{Gaia}
additionally provides low resolution BP-RP spectra and radial velocity
measurements of bright stars (\emph{Gaia} radial velocity magnitudes
\textless16) \citep{gaiacollaboration+2016}. For this work,
\emph{Gaia}'s most relevant measurement are \(G\) magnitude,
\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
  change in right ascension \(\alpha\) and declination \(\delta\),
  typically in units of mili-arcsecond (mas) per year. \(\mu_{\alpha*}\)
  allows where \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for
  projection effects in \(\alpha\).}

\emph{Gaia} has revolutionized many fields of astronomy including the
Milky Way and the Local Group. For example, 6D kinematic classification
of Milky Way stars has lead to the discovery of past mergers or Milky
Way building blocks like \emph{Gaia}-sausage Enceladus
\citep[e.g.,][]{helmi+2018}, out-of-equilibrium structures like the
\emph{Gaia} snail {[}e.g., \citet{antoja+2018l}, and dynamical effects
of the Milky Way's spiral arms and the bar in the solar neighbourhood
\citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
helped find and constrain numerous stellar streams
\citep{bonaca+price-whelan2025}. Altogether, \emph{Gaia} has revealed
the hierarchical formation and complex, evolving structure of our own
galaxy.

For Milky Way satellites, \emph{Gaia} has enabled well-constrained
orbital analysis and facilitated precise stellar membership
determination. Before, proper motions of dwarf galaxies were spase or
undetermined. Few galaxies had precisely measured proper motions, often
from Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
\emph{Gaia} allowed for the first systematic determinations of Milky Way
satellite proper motions, and of unprecedented precision
\citep{MV2020a, pace+li2019}. While the proper motion uncertainty on a
typical dwarf member star is often with large uncertainties, by
combining the proper motions for 100s to 1000s of stars in \emph{Gaia},
incredibly precise proper motion measurements can be determined,
sometimes only limited by \emph{Gaia}'s systematic error floor
\citep[e.g.,][]{MV2020a}. Proper motions have ushered in a new dynamical
era for MW satellites studies, where we often have precise
determinations of the dwarf galaxy's recent orbit under a given MW
potential. In addition, \emph{Gaia} helps seperate out contaiminating MW
foreground stars. By measuring parallaxes or proper motions
significantly different from the dwarf galaxy, many stars can be
immediately removed as non-members
\citep[e.g.,][\citet{jensen2024}]{battaglia+2022}.

@@

Today, we know the Milky Way system is teeming with satellites.
Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
including dwarf galaxies, ambiguous systems, and globular clusters.
Advances in telescopes and observations has accelerated progress on the
nature of dwarf galaxy and introduced new questions. Deep digital
photometric surveys have more than quadrupled the number of known Milky
Way dwarf galaxies \citep{simon2019}. Upcoming and ongoing surveys, like
the Vera Rubin Observatory's Legacy Survey for Space and Time, will
continue to probe fainter and fainter dwarf galaxies. In addition, large
aperture multi-object spectrographs have revealed the detailed and
complex inner chemodynamical structure of dwarf galaxies. Beyond precise
structural and kinematic properties of dwarf galaxies, modern
observations allow for the separation of multiple stellar populations,
detailed constraints on the dark matter density profiles, and hints of
tidal disruption or stellar halos.

Of local dwarfs, the classical systems still remain the best studied and
with most strongly constrained parameters. While extending 1-2 degrees
across the sky, making deriving properties trickier
\citep[e.g.,][]{mateo1998}, these dwarfs contain large numbers of bright
(giant) stars, allowing thousands of stars to be observed with deep
photometry and spectroscopy \citep[e.g.,][]{tolstoy+2023, pace+2020}. As
a result, the determination of fundamental properties such as the
position, size, orientation, distances, proper motions, line-of-sight
(LOS) velocities, and velocity dispersions \(\sigma_v\), are all
relatively well constrained and converged among different studies today.
However, ongoing research continues to redefine our understanding of the
detailed structure of Milky Way satellites, hinting that these objects
may be more complex than at first glance.

Observations of dwarf galaxies have been the origin of several disputes
or \emph{small-scale} problems in cosmology \citep[see
review][]{bullock+boylan-kolchin2017}. For example, the mismatch between
the expected number of dwarf galaxies from simulations and the observed
abundance was known as the \emph{missing satellites problem}.
Additionally, a number of observations showed that many dwarf galaxies,
but not all, possess dark matter ``cores'' as measured from circular
velocity curves \citep[e.g.,][]{moore1994, adams+2014, oh+2015} or dwarf
spheroidal \citep[albeit less constraining,
e.g.,][]{walker+penarrubia2011, read+walker+steger2019}. As a result,
numerous alternative forms of dark matter have been advocated as
solutions. \emph{Warm} dark matter, relativistic in the early universe
but cooler now, smooths out small-scale features and softens the cusps
of dark matter halos. \emph{Self-interacting} dark matter instead can
form cores but also the cores can collapse into a density peak.

However, these tensions have eased as a result of improved consideration
of baryonic physics. For instance, understanding observational
limitations and completeness has eased tension with the abundance of
dwarf galaxies \citep[e.g.,][]{kim+peter+hargis2018}. Also, recent
hydrodynamic simulations in particular have shown that strong feedback
can produce dark matter cores
\citep[e.g.,][]{tollet+2016, fitts+2017, orkney+2021}. Altogether, the
numerous past challenges for \(\Lambda\)CDM in the dwarf galaxy regime
illustrates the opportunity for dwarf galaxies to test the understanding
of galaxy formation and dark matter physics.

\subsection{Stellar density profiles}\label{stellar-density-profiles}

Surface density profiles efficiently characterize the shape of a galaxy.
At the most basic, fitting stellar densities provide properties such as
the shape, location, size, and orientation of a dwarf galaxy. However,
the details of a stellar density profile are essential for interpreting
the total mass of dwarf galaxies, their assembly and dynamical history,
and understanding correlations between dwarf galaxies
\citep[e.g.,][\citet{penarrubia+2009}, \citet{querci+2025},
\citet{lee+2018}]{herrmann+hunter+elmegreen2013}.

>>>>

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These little galaxies are typically
defined to be fainter than the LMC \citep[\(M_V \gtrsim -18\) or
similarly \(L \lesssim 10^9 L_\odot\),
e.g.,][]{hodge1971, mcconnachie2012}. As the smallest class of galaxies,
dwarfs are the most cosmologically numerous galaxies. Dwarf galaxies are
highly \emph{dark-matter dominated}, with mass to light ratios exceeding
100 to 1000 \(M_\odot/ L_\odot\). Because of their small total masses,
many dwarf galaxy satellites of the Milky Way are \emph{quenched}, with
little to no recent star formation. Milky Way dwarfs contain stellar
populations which are \emph{relics} from the early universe, consisting
of many of the oldest and most metal poor stars. Understanding the
properties of dwarf galaxies thus has implications across astronomy,
from cosmological structure formation on the smallest scale to extremely
metal-poor stellar populations to chemical evolution to galactic
dynamics. Of the classical dwarfs, we find that the Sculptor and Ursa
Minor dwarf spheroidal galaxies stand out with a more extended density
profile relative to an exponential, hinting at tidal effects. As a case
study in tidal effects and our understanding of dwarf galaxy formation
and evolution, we aim to understand the dynamical history of these
galaxies.

@@

@@

Dwarf galaxies have long raised conundrums in theories of galaxy
formation. The discovery of Fornax and Sculptor in 1938
\citep{shapley1938}\footnote{Technically, the Large and Small Magellanic
  Clouds (LMC, SMC) may be classified as dwarf galaxies, but these were
  likely always known to humans at southern latitudes.}, with no known
analogues, already presented such an enigma. Shapley presented these
dwarfs as a new type of \emph{stellar system} resembling the Magellanic
Clouds and globular clusters but did not attempt to speculate on the
exact nature. While dwarf galaxies were quickly understood to be
galaxies based on the inferred luminosities and sizes, their nature and
formation remained unclear for decades
\citep[e.g.,][]{hodge1971, gallagher+wyse1994}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial amounts of dark matter. From early determinations of the
velocity dispersion for the Sculptor and Ursa Minor dwarf spheroidal
(dSph) galaxies
\citep[e.g.,][\citet{aaronson+olszewski1987}]{aaronson1983}, inferred
mass-to-light ratios were at least 10 times larger than the values from
globular clusters scaled to the same sizes. While rather uncertain
initially, these values have corroborated with larger and more precise
samples \citep[e.g.,][]{hargreaves+1994}. Subsequently, several theories
attempting to understand the formation and observed properties of these
objects were proposed. Examples include: dwarf galaxies are undergoing
tidal dissolution resulting in extreme mass-to-light ratios
\citep[e.g.,][]{kuhn+miller1989}, presence of massive central black
holes \citep[e.g.,][]{strobel+lake1994}, the formation of dark
matter-free ``tidal dwarfs'' from past mergers
\citep[e.g.,][\citet{kroupa1997}]{lynden-bell1982}, or modified theories
of gravity \citep{milgrom1995}. However, consistency with CDM galaxy
formation was also not out of the question
\citep[e.g.,][]{dekel+silk1986}. Since then, we have known that dwarf
galaxies are among the darkest objects in the universe, and
understanding their properties is critical to understanding dark matter.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs} with magnitudes
\(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
and recent star formation. Fig.~\ref{fig:galaxy_images} shows the Large
Magellanic Cloud (LMC) as an example of an irregular, bright dwarf
galaxy displaying a bar. \textbf{Classical dwarfs} occupy intermediate
luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these
systems are old, gas-poor, and spheroidal. All Milky Way satellites
discovered before digital sky surveys are classicals. The 12 classical
dwarfs satellites of our Galaxy are Sagittarius, Fornax, Leo I,
Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
I, Sextans I, and Crater II.\footnote{While formally the dwarf galaxy
  names we discuss contain ``dwarf spheroidal'' (dSph), e.g.~Sculptor
  dSph, we omit this suffix for brevity.} The \textbf{ultra-faint}s
occupy the very faintest magnitudes (\(-7.7 \lesssim M_V\)). These
galaxies have minuscule stellar masses, tend to be more compact, and are
the darkest known galaxies. Altogether, dwarf galaxies span more than 15
orders in absolute magnitude.

Since Local Group dwarfs are nearby, we can study these galaxies on a
star-by-star basis. As a result, it is possible to measure the 3D
velocity and position of a star if we can measure the stars position,
line-of-sight (LOS) velocity, and proper motions. Unfortunately,
determining distances and full 3D velocities is challenging. The most
direct measurement of distance, parallax, requires precise tracking of a
star's sky position across a year. And while line-of-sight (LOS)
velocities are easily determined from spectroscopy, tangential
velocities, derived from proper motions and distances, are much more
challenging. Typically, measuring proper motions requires accurate (much
less than arcsecond) determinations of small changes in a star's
position over baselines of several years to decades. The full 6D
position and velocity information for stars has historically been known
for only a handful of stars.

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.png}
\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Fornax (DES
DR2), Sculptor (DES DR2), and Ursa Minor (UNWISE with Gaia point sources
over-plotted). Each image includes the galaxy's luminosity and a 200 pc
scale bar. \emph{TODO: Mark half-light radius with
circle?}}\label{fig:galaxy_images}
\end{figure}

@@

\emph{(db: these paragraphs start the same\ldots)}

\emph{Gaia} has redefined astrometry, providing photometry, positions,
proper motions, and parallaxes for over 1 billion stars
\citep{gaiacollaboration+2021}. Launched in 2013, \emph{Gaia} is a
space-based, all-sky survey telescope situated at the Sun-Earth L2
Lagrange point \citep{gaiacollaboration+2016}. While \emph{Gaia}
completed its space-based mission in 2025, two forthcoming data releases
remain. Determining absolute parallax measurement is facilitated by the
observation that stars in different regions of the sky are affected by
parallax motion with different phases. By imagining two regions
separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
tiny changes in relative positions of stars across small and large
angles. Combining measurements from multiple epochs across several
years, an absolute all-sky reference frame is derived from which
parallax and proper motions are derived. In addition to astrometry,
\emph{Gaia} measures photometry in the wide \emph{G} band (330-1050nm)
and colours from the blue photometer (BP, 330-680 nm) and red photometer
(RP, 640-1050 nm). \emph{Gaia} additionally provides low resolution
BP-RP spectra and radial velocity measurements of bright stars
(\emph{Gaia} radial velocity magnitudes \textless16)
\citep{gaiacollaboration+2016}. For this work, \emph{Gaia}'s most
relevant measurement are \(G\) magnitude, \(G_{\rm BP} - G_{\rm RP}\)
colour, \((\alpha, \delta)\) position, and
\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
  change in right ascension, \(\alpha\), and declination, \(\delta\),
  typically in units of mili-arcsecond (mas) per year.
  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
  effects in \(\alpha\).}

\emph{Gaia} has revolutionized many fields of astronomy including
studies of the Milky Way and the Local Group. The Local Group is defined
as galaxies which are approximately bound to the Milky Way-Andromeda
system, or within about 1 Mpc from the Milky Way \citep[e.g.,][ and
references therein]{mcconnachie2012}. For example, 6D kinematic
classification of Milky Way stars led to the discovery of past mergers
or Milky Way building blocks like \emph{Gaia}-Sausage Enceladus
\citep[e.g.,][]{helmi+2018}, out-of-equilibrium structures like the
\emph{Gaia} snail {[}e.g., \citet{antoja+2018}, and dynamical effects of
the Milky Way's spiral arms and the bar in the solar neighbourhood
\citep{hunt+vasiliev2025}. Moving to the Milky Way halo, \emph{Gaia} has
helped find and constrain numerous stellar streams
\citep{bonaca+price-whelan2025}. Altogether, \emph{Gaia} has revealed
the hierarchical formation and complex, evolving structure of our own
Galaxy.

For Milky Way satellites, \emph{Gaia} has enabled well-constrained
orbital analysis and facilitated precise stellar membership
determination. Before, proper motions of dwarf galaxies were sparse or
undetermined. Few galaxies had precisely measured proper motions, often
from Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
\emph{Gaia} allowed for the first systematic determinations of Milky Way
satellite proper motions of unprecedented precision
\citep{MV2020a, pace+li2019}. While the proper motion uncertainty on a
typical dwarf member star is often large, by combining the proper
motions for 100s to 1000s of stars in \emph{Gaia}, precise proper motion
measurements can be determined, sometimes only limited by \emph{Gaia}'s
systematic error floor \citep[e.g.,][]{MV2020a}. Proper motions have
ushered in a new dynamical era for MW satellite studies, where we can
derive precise orbits assuming a given MW potential. In addition,
\emph{Gaia} helps separate out contaminating MW foreground stars. By
measuring parallaxes or proper motions, many background and foreground
stars can be removed as non-members
\citep[e.g.,][\citet{jensen+2024}]{battaglia+2022}.

@@

Today, we know that the Milky Way system is teeming with satellites.
Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
including dwarf galaxies, ambiguous systems (whose nature as a
dwarf-like or cluster-like system remains uncertain), and globular
clusters. Advances in observational techniques has accelerated progress
in our understanding of dwarf galaxies and introduced new questions.
Deep digital photometric surveys have more than quadrupled the number of
known Milky Way dwarf galaxies \citep{simon2019}. Upcoming and ongoing
surveys, like the Vera Rubin Observatory's Legacy Survey for Space and
Time \citep{ivezic+2019}, will likely identify even fainter dwarf
galaxies. In addition, large aperture multi-object spectrographs have
revealed the detailed and complex inner chemodynamical structure of
dwarf galaxies. Beyond precise structural and kinematic properties of
dwarf galaxies, modern observations allow for the separation of multiple
stellar populations, detailed constraints on the dark matter density
profiles, and hints of tidal disruption or stellar halos.

Of local dwarfs, the classical systems still remain the most studied
galaxies with the best constrained parameters. Because many Milky Way
dwarfs extend 1-2 degrees across the sky, measurements of some of their
properties can be challenging \citep[e.g.,][]{mateo1998}. These dwarfs
also contain large numbers of bright (giant) stars which can be followed
up spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a
result, the determination of fundamental properties such as the
position, size, orientation, distances, proper motions, line-of-sight
(LOS) velocities, and velocity dispersions \(\sigma_v\), are all
relatively well constrained today. However, ongoing research continues
to redefine our understanding of the detailed structure of Milky Way
satellites, hinting that these objects may be more complex than expected
at first glance.

Observations of dwarf galaxies have been the origin of several disputes
or \emph{small-scale} problems for \(\Lambda\)CDM \citep[see
review][]{bullock+boylan-kolchin2017}. For example, the mismatch between
the expected number of dwarf galaxies from simulations and the observed
abundance was known as the \emph{missing satellites problem}.
Additionally, a number of observations suggested that some dwarf
galaxies, although not all, possess dark matter ``cores''
\citep[e.g.,][]{moore1994, adams+2014, oh+2015, walker+penarrubia2011, read+walker+steger2019}.\footnote{In
  detail, gas-phase rotation curves are better able to differentiate
  between cores and cusps, whereas stellar kinematics is less
  constraining.} As a result, alternative forms of dark matter have been
advocated as solutions, such as Warm and Self-Interacting Dark Matter.

However, these tensions have eased as a result of improved understanding
of baryonic physics. For example, recent hydrodynamic simulations in
particular have shown that strong feedback can produce dark matter cores
\citep[e.g.,][]{tollet+2016, fitts+2017, benitez-llambay+2019, orkney+2021}.
Altogether, the numerous past challenges for \(\Lambda\)CDM in the dwarf
galaxy regime illustrates the opportunity for dwarf galaxies to test the
understanding of galaxy formation and dark matter physics.

\subsection{Typical stellar density
profiles}\label{typical-stellar-density-profiles}

Projected density profiles efficiently characterize the shape of a
galaxy. At its most basic, density profiles provide properties such as
the shape, location, size, and orientation of a dwarf galaxy. In
addition, the details of a stellar density profile can help interpret
the total mass of dwarf galaxies, their assembly and dynamical history,
and to understand scaling laws between dwarf galaxy properties
\citep[e.g.,][\citet{penarrubia+2009}, \citet{querci+2025},
\citet{lee+2018}]{herrmann+hunter+elmegreen2013}. We note that for
resolved galaxies, these profiles are in terms of stellar count density
instead of surface magnitudes.
